---
title: "Patient Survival & Outcome Prediction"
subtitle: "Statistical Survival Analysis and Machine Learning on WHAS500"
author: "Ayodeji"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    theme: cosmo
    code-fold: true
    code-summary: "Show code"
    fig-width: 9
    fig-height: 6
    embed-resources: true
    smooth-scroll: true
execute:
  warning: false
  message: false
  echo: true
---

```{r setup, include=FALSE}
# Load R packages and results
library(survival)
library(ggplot2)
library(dplyr)
library(jsonlite)
library(knitr)

# Read results written by run_analysis.R
results_path <- file.path("data", "results", "survival_results.json")
if (file.exists(results_path)) {
  r_results <- fromJSON(results_path)
} else {
  warning("R results not found. Run: Rscript r/run_analysis.R")
  r_results <- NULL
}

cox_coef_path <- file.path("data", "results", "cox_coefficients.csv")
if (file.exists(cox_coef_path)) {
  cox_coef <- read.csv(cox_coef_path)
} else {
  cox_coef <- NULL
}
```

```{python setup-python, include=FALSE}
import sys, json, warnings
sys.path.insert(0, '.')
warnings.filterwarnings('ignore')

import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
import matplotlib
matplotlib.rcParams['figure.dpi'] = 120

from pathlib import Path

# Load ML results if available
ml_results_path = Path('data/results/ml_results_whas500.json')
ml_results = json.loads(ml_results_path.read_text()) if ml_results_path.exists() else {}
```

## Executive Summary

This report presents a comprehensive survival and outcome prediction analysis
of the **Worcester Heart Attack Study (WHAS500)** dataset — 500 patients
hospitalised with acute myocardial infarction between 1975 and 2001.

Two complementary analyses are presented:

1. **Survival analysis in R** — Kaplan-Meier curves, log-rank tests, and Cox
   proportional hazards modelling to understand time-to-event outcomes.

2. **Machine learning in Python** — Logistic Regression, Random Forest,
   XGBoost, and LightGBM compared for 30-day mortality prediction,
   with SHAP explainability.

::: {.callout-warning}
**Clinical disclaimer**: This analysis is for research and educational
purposes only.  The models described here have not been validated for
clinical use and must not be used to guide patient management.
:::

---

## 1. Dataset Overview

```{r dataset-overview}
if (!is.null(r_results)) {
  cat(sprintf("Patients        : %d\n",   r_results$n_patients))
  cat(sprintf("Deaths          : %d\n",   r_results$n_events))
  cat(sprintf("Mortality rate  : %.1f%%\n", r_results$event_rate_pct))
  cat(sprintf("Median survival : %d days\n",
    r_results$kaplan_meier$overall$median_survival_days))
  cat(sprintf("1-year survival : %.1f%%\n",
    r_results$kaplan_meier$overall$survival_1yr_pct))
  cat(sprintf("5-year survival : %.1f%%\n",
    r_results$kaplan_meier$overall$survival_5yr_pct))
}
```

```{python dataset-python}
from src.data.extract import load_whas500
from src.data.transform import clean_whas500, engineer_whas500_features

whas = clean_whas500(load_whas500())
whas = engineer_whas500_features(whas)

# Summary table
summary = whas[['age', 'bmi', 'sysbp', 'hr', 'los']].describe().T
summary.columns = ['N', 'Mean', 'SD', 'Min', 'Q25', 'Median', 'Q75', 'Max']
summary = summary[['Mean', 'SD', 'Median', 'Min', 'Max']].round(1)
print("Clinical feature summary (WHAS500):")
print(summary.to_string())
```

---

## 2. Survival Analysis

### 2.1 Kaplan-Meier Curves

The Kaplan-Meier estimator is a non-parametric method for estimating the
survival function $S(t) = P(T > t)$ — the probability of surviving beyond
time $t$.  No distributional assumptions are made.

```{r km-overall, fig.cap="Overall Kaplan-Meier survival curve for the WHAS500 cohort. Shaded area shows 95% confidence interval. Dashed vertical line marks median survival time."}
km_plot_path <- file.path("data", "results", "km_overall.png")
if (file.exists(km_plot_path)) {
  knitr::include_graphics(km_plot_path)
} else {
  cat("Run: Rscript r/run_analysis.R to generate plots")
}
```

```{r km-overall-interpretation, results='asis'}
if (!is.null(r_results)) {
  km <- r_results$kaplan_meier$overall
  cat(sprintf(
    "**Interpretation**: half of the cohort survived past %d days (~%.1f years) -- that is the median survival time. Survival falls fastest early on: only %.1f%% survive to one year, and it keeps declining more gradually after that, down to %.1f%% by five years. This front-loaded drop is typical for acute MI -- the highest risk of death is in the days and months immediately after the event, with a longer tail of later cardiovascular deaths.\n",
    km$median_survival_days, km$median_survival_days / 365.25,
    km$survival_1yr_pct, km$survival_5yr_pct
  ))
}
```

### 2.2 Survival by Age Group

The log-rank test evaluates whether survival curves differ significantly
between groups.  $H_0$: no difference in survival between age groups.

```{r km-age, fig.cap="Kaplan-Meier curves stratified by age group. Log-rank p-value shown on plot.", results='asis'}
km_age_path <- file.path("data", "results", "km_by_age.png")
if (file.exists(km_age_path)) knitr::include_graphics(km_age_path)

if (!is.null(r_results)) {
  lr <- r_results$kaplan_meier$log_rank_age
  cat(sprintf(
    "\nLog-rank test (age groups): χ² = %.3f, df = 2, p = %.4f — %s\n",
    lr$chisq,
    lr$p_value,
    ifelse(lr$significant, "SIGNIFICANT difference in survival", "no significant difference")
  ))
  cat(sprintf(
    "\n**Interpretation**: survival differs significantly across age groups (p %s). As the plot shows, the youngest group (<60) maintains the highest survival probability throughout follow-up, while the oldest group (>75) has the steepest, most sustained decline. Age keeps separating outcomes across the whole follow-up period, not just in the initial post-MI window.\n",
    ifelse(lr$p_value < 0.001, "< 0.001", sprintf("= %.4f", lr$p_value))
  ))
}
```

### 2.3 Survival by Sex

```{r km-sex, fig.cap="Kaplan-Meier curves stratified by sex.", results='asis'}
km_sex_path <- file.path("data", "results", "km_by_sex.png")
if (file.exists(km_sex_path)) knitr::include_graphics(km_sex_path)

if (!is.null(r_results)) {
  lr <- r_results$kaplan_meier$log_rank_sex
  cat(sprintf(
    "\nLog-rank test (sex): χ² = %.3f, p = %.4f — %s\n",
    lr$chisq, lr$p_value,
    ifelse(lr$significant, "SIGNIFICANT", "not significant")
  ))
  cat(sprintf(
    "\n**Interpretation**: survival differs significantly by sex in this unadjusted comparison (p = %.4f). On its own, this does not mean sex independently causes worse outcomes -- see the Cox model below, which adjusts for age and other clinical factors simultaneously and tells a more complete story.\n",
    lr$p_value
  ))
}
```

---

## 3. Cox Proportional Hazards Model

The Cox model estimates the **hazard ratio (HR)** for each feature:

- $HR > 1$: feature increases risk of death
- $HR < 1$: feature is protective (reduces risk)  
- $HR = 1$: no effect

$$h(t|X) = h_0(t) \exp(\beta_1 X_1 + \beta_2 X_2 + \cdots + \beta_p X_p)$$

```{r cox-results}
if (!is.null(r_results)) {
  cat(sprintf("Concordance index (C-statistic): %.3f (SE = %.3f)\n",
    r_results$cox_model$concordance,
    r_results$cox_model$concordance_se))
  cat(sprintf("AIC: %.1f\n\n", r_results$cox_model$aic))
  cat("Interpretation: C-index > 0.7 indicates good discrimination.\n")
}
```

```{r cox-table}
if (!is.null(cox_coef)) {
  display_cox <- cox_coef %>%
    select(feature, hazard_ratio, hr_lower, hr_upper, p.value, direction) %>%
    mutate(
      HR_CI = sprintf("%.3f (%.3f–%.3f)", hazard_ratio, hr_lower, hr_upper),
      p_val = ifelse(p.value < 0.001, "<0.001", sprintf("%.3f", p.value)),
      sig   = ifelse(p.value < 0.05, "*", "")
    ) %>%
    select(Feature=feature, `HR (95% CI)`=HR_CI, `p-value`=p_val, `*`=sig)

  knitr::kable(display_cox, caption = "Cox Model Hazard Ratios (* p < 0.05)")
}
```

```{r cox-forest, fig.cap="Forest plot of Cox model hazard ratios. Red = significant risk factor, green = significant protective factor, grey = not significant."}
forest_path <- file.path("data", "results", "cox_forest.png")
if (file.exists(forest_path)) knitr::include_graphics(forest_path)
```

```{r cox-forest-interpretation, results='asis'}
if (!is.null(cox_coef)) {
  sig  <- cox_coef %>% filter(significant)
  risk <- sig %>% filter(direction == "increases_risk") %>% arrange(desc(hazard_ratio))
  prot <- sig %>% filter(direction == "protective")
  sex_row <- cox_coef %>% filter(feature == "sex")

  cat("**Interpretation**:\n\n")
  for (i in seq_len(nrow(risk))) {
    cat(sprintf(
      "- **%s**: HR = %.3f (95%% CI %.3f–%.3f, p %s) — each one-unit increase raises the hazard of death by %.1f%%.\n",
      risk$feature[i], risk$hazard_ratio[i], risk$hr_lower[i], risk$hr_upper[i],
      ifelse(risk$p.value[i] < 0.001, "< 0.001", sprintf("= %.3f", risk$p.value[i])),
      (risk$hazard_ratio[i] - 1) * 100
    ))
  }
  for (i in seq_len(nrow(prot))) {
    cat(sprintf(
      "- **%s**: HR = %.3f (95%% CI %.3f–%.3f, p %s) — each one-unit increase *lowers* the hazard by %.1f%%.\n",
      prot$feature[i], prot$hazard_ratio[i], prot$hr_lower[i], prot$hr_upper[i],
      ifelse(prot$p.value[i] < 0.001, "< 0.001", sprintf("= %.3f", prot$p.value[i])),
      (1 - prot$hazard_ratio[i]) * 100
    ))
  }

  if (nrow(sex_row) == 1) {
    cat(sprintf(
      "\nNotably, **sex** (HR = %.3f, p = %.3f) is *not* significant here, even though the unadjusted Kaplan-Meier comparison above was (p = 0.0052). Once age and the other clinical factors are accounted for, the raw sex difference in survival looks like it is substantially explained by confounding — e.g. women in this cohort trending older on average — rather than sex independently driving risk. This is a good example of why an unadjusted comparison and a multivariate model can disagree, and why both are shown here rather than just one.\n",
      sex_row$hazard_ratio[1], sex_row$p.value[1]
    ))
  }
}
```

---

## 4. Machine Learning — Mortality Prediction

Four classifiers were trained on WHAS500 to predict 30-day mortality.
All models used stratified 5-fold cross-validation for evaluation.

```{python ml-comparison}
if ml_results:
    rows = []
    for name, res in ml_results.items():
        tm = res.get('test_metrics', {})
        rows.append({
            'Model':           name.replace('_', ' ').title(),
            'CV AUC':          f"{res['cv_metrics']['cv_roc_auc_mean']:.4f} ± {res['cv_metrics']['cv_roc_auc_std']:.4f}",
            'Test AUC':        f"{tm.get('roc_auc', 0):.4f}",
            'Avg Precision':   f"{tm.get('average_precision', 0):.4f}",
            'F1':              f"{tm.get('f1', 0):.4f}",
            'Recall':          f"{tm.get('recall', 0):.4f}",
            'Brier Score':     f"{tm.get('brier_score', 0):.4f}",
        })
    comparison_df = pd.DataFrame(rows).sort_values('Test AUC', ascending=False)
    print(comparison_df.to_string(index=False))
else:
    print("Run: predict train --dataset whas500")
```

---

## 5. SHAP Explainability

SHAP (SHapley Additive exPlanations) quantifies how much each feature
contributed to a specific prediction.

$$f(x) = \phi_0 + \sum_{i=1}^{p} \phi_i$$

where $\phi_0$ is the baseline (mean prediction) and $\phi_i$ is the
SHAP value for feature $i$ — how much that feature pushed the prediction
above or below the baseline.

```{python shap-global}
shap_path = Path('data/results/shap_whas500_logistic_regression.json')
if shap_path.exists():
    shap_res = json.loads(shap_path.read_text())
    print("Top 10 features by mean |SHAP| (Logistic Regression on WHAS500):\n")
    for feat in shap_res['global_importance'][:10]:
        bar = '#' * int(feat['mean_abs_shap'] * 200)
        print(f"  {feat['rank']:2d}. {feat['feature']:<22} {feat['mean_abs_shap']:.4f}  {bar}")
else:
    print("Run SHAP analysis first: python -m src.models.shap_explainer")
```

---

## 6. Clinical Interpretation

Based on the survival analysis and ML results:

**Key mortality predictors identified**:

1. **Age** — the strongest predictor in both the Cox model (HR > 1) and SHAP analysis.
   Each additional decade of age substantially increases mortality risk.

2. **Heart failure (CHF)** — cardiogenic shock and CHF complications
   approximately double the hazard of death (Cox HR ~2.0–2.5).

3. **Heart rate at admission** — elevated heart rate (tachycardia >100 bpm)
   is associated with significantly worse outcomes.

4. **Pulse pressure** (systolic − diastolic BP) — elevated pulse pressure
   is an independent predictor of cardiovascular mortality, consistent
   with its role as a marker of arterial stiffness.

5. **Sex** — female patients showed significantly different survival curves
   (log-rank p < 0.05), consistent with known sex differences in MI outcomes.

**Model performance**: Logistic Regression achieved the highest test AUC
(~0.88) on WHAS500, narrowly ahead of Random Forest (~0.88) and above
XGBoost/LightGBM (~0.85) — plausible on a dataset this size (500 patients),
where the relationships are close to linear once pulse_pressure and the
other engineered features are in play. Being the best-performing model
here is also convenient: Logistic Regression's coefficients are directly
interpretable, which matters for a clinical-facing tool.
The Cox C-index (~0.76) indicates good discrimination for a clinical model.

::: {.callout-note}
These findings are consistent with established clinical knowledge about
acute MI prognosis and serve to validate the modelling approach.
A prospective clinical study would be required to confirm these associations
in a new patient population.
:::

---

## 7. Methods Summary

| Component | Tool | Purpose |
|---|---|---|
| Data | WHAS500 (n=500) | Acute MI cohort, real survival times |
| Survival analysis | R 4.4 (survival, survminer) | KM curves, Cox PH model |
| ML training | Python 3.10 (sklearn, XGBoost, LightGBM) | Mortality classification |
| Explainability | SHAP (TreeExplainer) | Feature attribution |
| Validation | 5-fold stratified CV + held-out test set | Performance estimation |
| Report | Quarto | Reproducible R + Python document |

---

*Report generated with Quarto. All code is available in the project repository.*
*Run `quarto render report.qmd` to regenerate.*
